---
title: "class11: AlphaFold2 analysis"
author: "Ebony Michelle Argaez (PID: )"
format: pdf
---


Post process and inspecting modeling results from AlphaFold2

Results from AF Live in the folder/directory 'hivprdimer_23119'

```{r}
library(bio3d)

results_dir <- 'hivprdimer_23119_1/'

pdb_files <- list.files(results_dir, pattern = ".pdb", full.names=T)
```


Align and supperpose PDB models by using 'pdbaln()' 

```{r}
pdbs <- pdbaln(pdb_files, fit=T, exefile="msa")
```

## RMSD Matrix

A common measure of structural dis-similarity is called RMSD (root mean square distance)

```{r}
rd <- rmsd(pdbs)
rd
```

```{r}
library(pheatmap)

rownames(rd) <- paste0("m", 1:5)
colnames(rd) <- paste0("m", 1:5)


pheatmap(rd)
```

viewing them in Mol*, need to get fitted coords

```{r}
xyz <- pdbfit(pdbs, outpath="fitted")
```


A full atom based fitting or supperposition did not work very well because we have a multiple chains that are in different conformations.

Want to focus our supperposition on the most invarient part (the rigid "core" if you will)

```{r}
core <- core.find(pdbs)
core.inds <- core

```

```{r}
xyz <- pdbfit(pdbs, inds=core.inds, outpath="core_fitted")
```


to evaluate how good multi-chain or multi-domain models are we need to look at PAE scores (predicted aligned error) 

there are output as JSON format files. 

```{r}
pae_files <- list.files(results_dir, pattern = "0.json", full.names = T)
pae_files
```

```{r}
library(jsonlite)

pae1 <- read_json(pae_files[1], simplifyVector = T)
pae5 <- read_json(pae_files[5], simplifyVector = T)
```

```{r}
attributes(pae1)
```

```{r}
pae1$max_pae
pae5$max_pae
```

```{r}
plot.dmat(pae5$pae,
          xlab="Residue No.",
          ylab="Residue No.", 
          zlim=c(0,30)
          )

```


```{r}
plot.dmat(pae1$pae,
          xlab="Residue No.",
          ylab="Residue No.", 
          zlim=c(0,30)
          )
```

## Main points

-run AlphaFold on google compute infastructure ^_^
-read these results into R and process to help us make sense of these models and their PAE and pLDDT scores. 

