---
title: "lab08: Halloween mini project"
author: "Ebony Michelle Argaez (PID: A59026556)"
format: pdf
editor: visual
---

## Importing candy data
```{r}
candy <- read.csv("candy-data.csv", row.names = 1)


head(candy)
```

>Q1. How many different candy types are in this dataset?

```{r}
dim(candy)
```

85 candies


>Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

38 fruity candy types


## What is your favorite candy?

```{r}
candy["Twix", ]$winpercent
```

>Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

Swedish Fish my fav candy
```{r}
candy["Swedish Fish", ]$winpercent
```

winpercent Swedish Fish: 54.86111


>Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```
winpercent of kit kat: 76.7686

>Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

Winpercent Tootsie Roll Snack Bars :  49.6535

```{r}
library("skimr")
skim(candy)
```










>Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Winpercent is the variable that looks to be on a different scale. 

>Q7. What do you think a zero and one represent for the candy$chocolate column?

It tells you whether it's chocolate or not. 


>Q8. Plot a histogram of winpercent values

```{r}
hist(candy$winpercent)
```


>Q9. Is the distribution of winpercent values symmetrical?

No the distribution of winpercent is not symmetrical. 

>Q10. Is the center of the distribution above or below 50%?

Below 50%

>Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
choc.ind<- as.logical(candy$chocolate)
fruit.ind <- as.logical(candy$fruity)

choc.win <- candy[choc.ind,]$winpercent
fruit.win <- candy[fruit.ind,]$winpercent
```

```{r}
mean(choc.win)
mean(fruit.win)
```

>Q12. Is this difference statistically significant?

```{r}
t.test(choc.win, fruit.win)
```

Yes it'ssignificant because pvalue ( 2.871e-08) is less than 0.05.


## Overall Candy Rankins
>Q13. What are the five least liked candy types in this set?

```{r}
#installing tidyverse
library(tidyverse)
```

```{r}
#using base R
head(candy[order(candy$winpercent),], n=5)
```

```{r}
#using dplyr
candy %>% arrange(winpercent) %>% head(5)
```

The five least candy are Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbusters. 



>Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
candy %>% arrange(winpercent) %>% tail(5)

```


Top 5 candy: Snickers, Kitkat, Twix, Reese's miniatures, and Reese's Peanut Butter Cup. 

>Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)
#generating plot
ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```

>Q16 This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
#reordering plot
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col()
```

```{r}
#adding color to plot
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col(fill=my_cols)
```

>Q17. What is the worst ranked chocolate candy?

Worst: Sixlets

>Q18. What is the best ranked fruity candy?

Best: Starburst

## Taking a look at pricepercent

```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 15)

```

>Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Tootsie Roll Midgies

>Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )

```

Most expensive: Nik L Nip, Nestle Smarties, Ring Pop, Hershey's Krackel, Hershey's Milk Chocolate

## Exploring the correlation structure

```{r}
library(corrplot)
```

```{r}
cij <- cor(candy)
corrplot(cij)
```

>Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Fruity and chocolate

>Q23. Similarly, what two variables are most positively correlated?

chocolate and winpercent

## Principal Component Analysis

```{r}
pca <- prcomp(candy, scale=T)
summary(pca)
```

```{r}
plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab="PC2")

```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
library(plotly)
```

```{r, eval=FALSE}
ggplotly(p)
```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")

```

>Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

pluribus, hard, and fruity are variables picked up strongly by PC1 in the positive direction. Fruity is the strongest one. This makes sense because fruity and chocolate are the most popular ones. 
