---
title: "lab6supplemental"
author: "Ebony Michelle Argaez (PID: A96056556)"
format: pdf
editor: visual
---

> B. Next improve the below example code for the analysis of protein drug interactions by abstracting the main activities in your own new function. Then answer questions 1 to 6 below. It is recommended that you start a new Project in RStudio in a new directory andthen install the bio3d package noted in the R code below (N.B. you can use the command install.packages("bio3d") or the RStudio interface to do this).
Then run through the code to see if it works, fix any copy/paste errors before simplifyingto a core working code snippet, reducing any calculation duplication, and finally
transferring it into a more useful function for you.


```{r}
library(bio3d)
s1 <- read.pdb("4AKE") # kinase with drug
s2 <- read.pdb("1AKE") # kinase no drug
s3 <- read.pdb("1E4Y") # kinase with drug
s1.chainA <- trim.pdb(s1, chain="A", elety="CA")
s2.chainA <- trim.pdb(s2, chain="A", elety="CA")
s3.chainA <- trim.pdb(s1, chain="A", elety="CA")
s1.b <- s1.chainA$atom$b
s2.b <- s2.chainA$atom$b
s3.b <- s3.chainA$atom$b
plotb3(s1.b, sse=s1.chainA, typ="l", ylab="Bfactor")
plotb3(s2.b, sse=s2.chainA, typ="l", ylab="Bfactor")
plotb3(s3.b, sse=s3.chainA, typ="l", ylab="Bfactor")
```

 > Q1. What type of object is returned from the read.pdb() function?
 
This allows you to read a PDB coordinate file from the RCSB online database. The output is a character vector and a summary of the protein. 

> Q2. What does the 'trim.pdb()' function do?

This function results in a subset of the PDB structure previously chosen.

>Q3. What input parameter would turn off the marginal black and grey rectangles in plots and what do they represent in this case?

You can change sse=FALSE. The rectangles represent levels of secondary structures. 

>Q4. What would be a better plot to compare across the different proteins?

Not sure. Stacking scatter plots above each other to compare b-factors.

>Q5. Which proteins are more similar to each other in their B-factor trends. How could you quantify this?

```{r}
hc <- hclust( dist( rbind(s1.b, s2.b, s3.b) ) )
plot(hc)
```

The s1.b and s3.b proteins are more similar. 

>Q.6 How would you generalize the original code above to work with any set of inputprotein structures?

```{r}
#' Function to analyse protein drug interactions
#'
#' @param x results in 
#'
#' @return results in a scatter plot of Bfactors values 
#' @export
#'
#' @examples
#' if chosen protein is PDB:"4AKE"
#' p_struc("4AKE")

p_struc <- function (x) {
  #read a PDB file from the RCSB online database, save result
  s <- read.pdb(x) 
  #trim PDB based on chain and elety and save result
  s.chainA <- trim.pdb(s, chain = "A", elety = "CA")
  #get data from atomic coordinates and Bfactor
  s.b <- s.chainA$atom$b 
  #plots scatter plot of B-factor values 
  plotb3(s.b, sse = s.chainA, typ="l", ylab = "Bfactor")
}
```

Check to see if function works
```{r}
p_struc("4AKE")
p_struc("1AKE")
p_struc("1E4Y")

```

